\TheTitle

### 2019-11-25 Paper1/framework.tex
local convergence comment
Paper Jakob added
last comments from Martin

When a consistent numerical flux $\bm g:\mathbb{R}^m\times\mathbb{R}^m\to\mathbb{R}^m$, $\bm g = \bm g(\bm u_\ell, \bm u_r)$ is available for the original problem \eqref{eq:hyperbolicProblem}, then for the IPM system we can simply take the numerical flux
\begin{align*}
 \bm{\tilde G}(\bm{\hat u}_{j}^n,\bm{\hat u}_{j+1}^{n}) = \langle \bm g(\mathcal{U}(\bm{\hat u}_j^n),\mathcal{U}(\bm{\hat u}_{j+1}^n))\bm{\varphi}^T\rangle^T
\end{align*}
in \eqref{eq:IPMDiscretization}. Commonly, this integral cannot be evaluated analytically and therefore needs to be approximated by a quadrature rule
\begin{align*}
\langle h \rangle \approx \langle h \rangle_{Q} := \sum_{k=1}^Q w_k h(\bm{\xi}_k)f_{\Xi}(\bm{\xi}_k).
\end{align*}
The approximated numerical flux then becomes
\begin{align}\label{eq:numericalFluxIPM}
 \bm{G}(\bm{\hat u}_{j}^n,\bm{\hat u}_{j+1}^{n}) = \langle \bm g(\mathcal{U}(\bm{\hat u}_j^n),\mathcal{U}(\bm{\hat u}_{j+1}^n))\bm{\varphi}^T\rangle^T_Q.
\end{align}
Note that the numerical flux requires evaluating the ansatz $\mathcal{U}(\bm{\hat u}_j^n)$. To simplify notation, we define $\bm{u}_{s}:\mathbb{R}^p \to \mathbb{R}^p$,
\begin{align*}
\bm{u}_{s}(\bm\Lambda):=\left( \nabla_{\bm{u}} s \right)^{-1}(\bm\Lambda),
\end{align*}
meaning that the IPM ansatz \eqref{eq:ansatz} at cell $j$ in timestep $n$ can be written as
\begin{align*}
\mathcal{U}(\bm{\hat u}_j^n) = \bm{u}_{s}(\bm{\hat{\lambda}}(\bm{\hat u}_j^n)^T \bm{\varphi}).
\end{align*}
The computation of the dual variables $\bm{\hat\lambda}_j^n:=\bm{\hat\lambda}(\bm{\hat u}_j^n)$ requires solving the dual problem \eqref{eq:dualProblem} for the moment vector $\bm{\hat u}_{j}^{n}$. Hence, to determine the dual variables for a given moment vector $\bm{\hat{u}}$, the cost function
\begin{align}\label{eq:L}
L(\bm{\lambda};\bm{\hat{u}}) := \langle s_*(\bm{\lambda}^T \bm\varphi)\rangle_Q - \sum_{i\leq M}\bm{\lambda}_i^T \bm{\hat u}_i
\end{align}
needs to be minimized. Hence, one needs to find the root of
\begin{align*}
\nabla_{\bm{\lambda_i}}L(\bm{\lambda};\bm{\hat{u}}) = \langle \nabla s_*(\bm{\lambda}^T \bm\varphi)\bm\varphi^T\rangle_Q^T - \bm{\hat u}_i = \langle \bm u_s(\bm{\lambda}^T \bm\varphi)\bm\varphi^T\rangle_Q^T - \bm{\hat u}_i,
\end{align*}
where we used $\nabla s_* \equiv \bm u_s$. The root is usually determined by using Newton's method. For simplicity, let us define the full gradient of the Lagrangian to be $\nabla_{\bm{\lambda}}L(\bm{\lambda};\bm{\hat{u}})\in\mathbb{R}^{N\cdot m}$, i.e. we store all entries in a vector. Newton's method uses the iteration function $\bm{d}:\mathbb{R}^{N\times m}\times\mathbb{R}^{N\times m}\to\mathbb{R}^{N\times m}$,
\begin{align}\label{eq:dualIterationFunction}
\bm{d}(\bm{\lambda},\bm{\hat{u}}):= \bm{\lambda}-\bm{H}(\bm{\lambda})^{-1}\cdot\nabla_{\bm{\lambda}}L(\bm{\lambda};\bm{\hat{u}}),
\end{align}
where $\bm H\in\mathbb{R}^{m \cdot  N\times m\cdot N}$ is the Hessian of \eqref{eq:L}, given by
\begin{align*}
\bm{H}(\bm{\lambda}) := \langle \nabla \bm{u}_{s} (\bm{\lambda}^T\bm{\varphi})\otimes\bm{\varphi}\bm{\varphi}^T\rangle_Q^{T}.
\end{align*}
The function $\bm d$ will in the following be called dual iteration function. Now, the Newton iteration for spatial cell $j$ is given by
\begin{align}\label{eq:dualIteration1}
\bm{\lambda}^{(l+1)}_j = \bm{d}(\bm{\lambda}_j^{(l)},\bm{\hat{u}_j}).
\end{align}
The exact dual state is then obtained by computing the fixed point of $\bm{d}$, meaning that one converges the iteration \eqref{eq:dualIteration1}, i.e. $\bm{\hat\lambda}_j^n:=\bm{\hat\lambda}(\bm{\hat u_j^n})=\lim_{m\rightarrow\infty}\bm{d}(\bm{\lambda}_j^{(l)},\bm{\hat{u}_j^n})$.
To obtain a finite number of iterations for the iteration in cell $j$, a stopping criterion 
\begin{align}\label{eq:tauCrit}
\sum_{i=0}^m\left\Vert \nabla_{\bm{\lambda_i}}L(\bm{\lambda}_j^{(l)};\bm{\hat{u}}_j^n) \right\Vert < \tau
\end{align}
is used.

@@

Intrusive numerical methods which compute arising integrals analytically and therefore directly depend on the moments (i.e. they do not necessitate the evaluation of the gPC expansion on quadrature points) can be constructed by performing a gPC expansion on the system flux directly \cite{debusschere2004numerical}. Examples can be found in \cite{hu2015stochastic,hu2016stochastic,tryoen2010instrusive}. While the analytic computation of arising integrals is not always more efficient \cite[Section 6]{ghanem1998stochastic}, it can also complicate recycling a deterministic solver. See \ref{app:costNumFlux} for a comparison of numerical costs when using Burgers' equation. However, when not using a quadratic entropy in the IPM method or when the physical flux of the deterministic problem is not a polynomial, it is not clear how many quadrature points the numerical quadrature rule requires to guarantee a sufficiently small quadrature error. We will study the approximation properties of IPM with different quadrature orders in Section~\ref{sec:resultsNACA1D}.

>>>>

When a consistent numerical flux $\bm g:\mathbb{R}^m\times\mathbb{R}^m\to\mathbb{R}^m$, $\bm g = \bm g(\bm u_\ell, \bm u_r)$ is available for the original problem \eqref{eq:hyperbolicProblem}, then for the IPM system we can simply take the numerical flux
\begin{align*}
 \bm{\tilde G}(\bm{\hat u}_{j}^n,\bm{\hat u}_{j+1}^{n}) = \langle \bm g(\mathcal{U}(\bm{\hat u}_j^n),\mathcal{U}(\bm{\hat u}_{j+1}^n))\bm{\varphi}^T\rangle^T
\end{align*}
in \eqref{eq:IPMDiscretization}. Commonly, this integral cannot be evaluated analytically and therefore needs to be approximated by a quadrature rule
\begin{align*}
\langle h \rangle \approx \langle h \rangle_{Q} := \sum_{k=1}^Q w_k h(\bm{\xi}_k)f_{\Xi}(\bm{\xi}_k).
\end{align*}
The approximated numerical flux then becomes
\begin{align}\label{eq:numericalFluxIPM}
 \bm{G}(\bm{\hat u}_{j}^n,\bm{\hat u}_{j+1}^{n}) = \langle \bm g(\mathcal{U}(\bm{\hat u}_j^n),\mathcal{U}(\bm{\hat u}_{j+1}^n))\bm{\varphi}^T\rangle^T_Q.
\end{align}
Note that the numerical flux requires evaluating the ansatz $\mathcal{U}(\bm{\hat u}_j^n)$. To simplify notation, we define $\bm{u}_{s}:\mathbb{R}^p \to \mathbb{R}^p$,
\begin{align*}
\bm{u}_{s}(\bm\Lambda):=\left( \nabla_{\bm{u}} s \right)^{-1}(\bm\Lambda),
\end{align*}
meaning that the IPM ansatz \eqref{eq:ansatz} at cell $j$ in timestep $n$ can be written as
\begin{align*}
\mathcal{U}(\bm{\hat u}_j^n) = \bm{u}_{s}(\bm{\hat{\lambda}}(\bm{\hat u}_j^n)^T \bm{\varphi}).
\end{align*}
The computation of the dual variables $\bm{\hat\lambda}_j^n:=\bm{\hat\lambda}(\bm{\hat u}_j^n)$ requires solving the dual problem \eqref{eq:dualProblem} for the moment vector $\bm{\hat u}_{j}^{n}$. Hence, to determine the dual variables for a given moment vector $\bm{\hat{u}}$, the cost function
\begin{align}\label{eq:L}
L(\bm{\lambda};\bm{\hat{u}}) := \langle s_*(\bm{\lambda}^T \bm\varphi)\rangle_Q - \sum_{i\leq M}\bm{\lambda}_i^T \bm{\hat u}_i
\end{align}
needs to be minimized. Hence, one needs to find the root of
\begin{align*}
\nabla_{\bm{\lambda_i}}L(\bm{\lambda};\bm{\hat{u}}) = \langle \nabla s_*(\bm{\lambda}^T \bm\varphi)\bm\varphi^T\rangle_Q^T - \bm{\hat u}_i = \langle \bm u_s(\bm{\lambda}^T \bm\varphi)\bm\varphi^T\rangle_Q^T - \bm{\hat u}_i,
\end{align*}
where we used $\nabla s_* \equiv \bm u_s$. The root is usually determined by using Newton's method. For simplicity, let us define the full gradient of the Lagrangian to be $\nabla_{\bm{\lambda}}L(\bm{\lambda};\bm{\hat{u}})\in\mathbb{R}^{N\cdot m}$, i.e. we store all entries in a vector. Newton's method uses the iteration function $\bm{d}:\mathbb{R}^{N\times m}\times\mathbb{R}^{N\times m}\to\mathbb{R}^{N\times m}$,
\begin{align}\label{eq:dualIterationFunction}
\bm{d}(\bm{\lambda},\bm{\hat{u}}):= \bm{\lambda}-\bm{H}(\bm{\lambda})^{-1}\cdot\nabla_{\bm{\lambda}}L(\bm{\lambda};\bm{\hat{u}}),
\end{align}
where $\bm H\in\mathbb{R}^{m \cdot  N\times m\cdot N}$ is the Hessian of \eqref{eq:L}, given by
\begin{align*}
\bm{H}(\bm{\lambda}) := \langle \nabla \bm{u}_{s} (\bm{\lambda}^T\bm{\varphi})\otimes\bm{\varphi}\bm{\varphi}^T\rangle_Q^{T}.
\end{align*}
%Inside the Newton update \eqref{eq:dualIterationFunction}, we abuse notation for better readability by making use of
%\begin{align*}
%\left(\bm{H}^{-1}\cdot\nabla_{\bm{\lambda}}L\right)_{ij} := \sum_{i' = 1}^N\sum_{j' = 1}^m \left(\bm{H}^{-1}\right)_{m(j-1) + i,m(j'-1) + i'}\cdot\nabla_{\bm{\lambda}}L_{i'j'}.
%\end{align*}
The function $\bm d$ will in the following be called dual iteration function. Now, the Newton iteration for spatial cell $j$ is given by
\begin{align}\label{eq:dualIteration1}
\bm{\lambda}^{(l+1)}_j = \bm{d}(\bm{\lambda}_j^{(l)},\bm{\hat{u}_j}).
\end{align}
The exact dual state is then obtained by computing the fixed point of $\bm{d}$, meaning that one converges the iteration \eqref{eq:dualIteration1}, i.e. $\bm{\hat\lambda}_j^n:=\bm{\hat\lambda}(\bm{\hat u_j^n})=\lim_{l\rightarrow\infty}\bm{d}(\bm{\lambda}_j^{(l)},\bm{\hat{u}_j^n})$.
To obtain a finite number of iterations for the iteration in cell $j$, a stopping criterion 
\begin{align}\label{eq:tauCrit}
\sum_{i=0}^m\left\Vert \nabla_{\bm{\lambda_i}}L(\bm{\lambda}_j^{(l)};\bm{\hat{u}}_j^n) \right\Vert < \tau
\end{align}
is used.

@@

Intrusive numerical methods which compute arising integrals analytically and therefore directly depend on the moments (i.e. they do not necessitate the evaluation of the gPC expansion on quadrature points) can be constructed by performing a gPC expansion on the system flux directly \cite{debusschere2004numerical}. Examples can be found in \cite{hu2015stochastic,hu2016stochastic,tryoen2010instrusive,durrwachterahigh} for the computation of numerical fluxes and sources. While the analytic computation of arising integrals is not always more efficient \cite[Section 6]{ghanem1998stochastic}, it can also complicate recycling a deterministic solver. See \ref{app:costNumFlux} for a comparison of numerical costs when using Burgers' equation. However, when not using a quadratic entropy in the IPM method or when the physical flux of the deterministic problem is not a polynomial, it is not clear how many quadrature points the numerical quadrature rule requires to guarantee a sufficiently small quadrature error. We will study the approximation properties of IPM with different quadrature orders in Section~\ref{sec:resultsNACA1D}.

### 2019-11-25 Paper1/implementation.tex
continued
last comments from Martin

\section{Parallelization and Implementation}
\label{sec:parallel}
It remains to discuss the parallelization of the presented algorithms. In order to minimize the parallelization overhead, our goal is to minimize the need to send data between processors. Note that the dual problem (line 8 in Algorithm~\ref{alg:ad-IPM} and line 5 in Algorithm~\ref{alg:adosIPM}) does not require communication, i.e. it suffices to distribute the spatial cells between processors. In contrast to that, the finite volume update (line 14 in Algorithm~\ref{alg:ad-IPM} and line 8 in Algorithm~\ref{alg:adosIPM}) requires communication, since values at neighboring cells need to be evaluated. Hence, distributing the spatial mesh between processors will yield communication overhead since data needs to be sent whenever a stencil cell lies on a different processor. Therefore, we choose to parallelize the quadrature points, which again minimizes communication effort. As mentioned in \ref{app:costNumFlux}, we first compute the solution at stencil cells for all quadrature points. I.e. we determine $\bm u^{(j)}_k\in\mathbb{R}^p$ and the corresponding stencil cells for $k = 1,\cdots,Q$ by
\begin{align*}
\bm u^{(j-1)}_k := \bm u_{s}(\bm{\lambda}_{j-1}^T\bm{\varphi}_{\ell_1'}(\bm \xi_k)), \enskip \bm u^{(j)}_k := \bm u_{s}(\bm{\lambda}_{j}^T\bm{\varphi}_{\ell_2'}(\bm \xi_k)), \enskip \bm u^{(j+1)}_k := \bm u_{s}(\bm{\lambda}_{j+1}^T\bm{\varphi}_{\ell_3'}(\bm \xi_k)).
\end{align*}
Then, the finite volume update function \eqref{eq:adaptiveFVUpdate} can be written as
\begin{align}\label{eq:momentUpQuadrature}
\bm{c}_{\ell}^{\bm{\ell}'}&\left(\bm{\lambda}_{1},\bm{\lambda}_2,\bm{\lambda}_3\right)=\sum_{k=1}^Q w_k \left[\bm u^{(j)}_k- \frac{\Delta t}{\Delta x}\left(\bm g( \bm u^{(j)}_k,\bm u^{(j+1)}_k )- \bm g( \bm u^{(j-1)}_k,\bm u^{(j)}_k )\right)\right]\bm{\varphi}_{\ell}(\bm \xi_k)^T.
\end{align}
Instead of distributing the spatial mesh on the different processors, we now distribute the quadrature set, i.e. the sum in \eqref{eq:momentUpQuadrature} can be computed in parallel. Now after having performed the dual update, the dual variables are send to all processors. With these variables, each processor computes the solution on its portion of the quadrature set and then computed its part of the sum in \eqref{eq:momentUpQuadrature} on all spacial cells. All parts from the different processors are then added together and the full time-updated moments are distributed to all processors. From here, the dual update can again be performed. The standard IPM Algorithm~\ref{alg:IPM} and One-Shot IPM Algorithm~\ref{alg:osIPM} use this parallelization strategy accordingly. Again, we point out that stochastic-Galerkin is a variant of IPM, i.e. all presented techniques for IPM can also be used for SG. The SC algorithm that we use to compare intrusive with non-intrusive methods uses a given deterministic solver as a black box. Here, we distribute the quadrature set between all processors. Note that both, SC and IPM are based on the same deterministic solver, i.e. we use the same deterministic numerical flux $\bm g$. This allows a fair comparison of the different intrusive and non-intrusive techniques.

>>>>

\section{Parallelization and Implementation}
\label{sec:parallel}
It remains to discuss the parallelization of the presented algorithms. In order to minimize the parallelization overhead, our goal is to minimize the need to send data between processors. Note that the dual problem (line 8 in Algorithm~\ref{alg:ad-IPM} and line 5 in Algorithm~\ref{alg:adosIPM}) does not require communication, i.e. it suffices to distribute the spatial cells between processors. In contrast to that, the finite volume update (line 14 in Algorithm~\ref{alg:ad-IPM} and line 8 in Algorithm~\ref{alg:adosIPM}) requires communication, since values at neighboring cells need to be evaluated. Hence, distributing the spatial mesh between processors will yield communication overhead since data needs to be sent whenever a stencil cell lies on a different processor. Therefore, we choose to parallelize the quadrature points, which again minimizes communication effort. As mentioned in \ref{app:costNumFlux}, we first compute the solution at stencil cells for all quadrature points. I.e. we determine $\bm u^{(j)}_k\in\mathbb{R}^m$ and the corresponding stencil cells for $k = 1,\cdots,Q$ by
\begin{align*}
\bm u^{(j-1)}_k := \bm u_{s}(\bm{\lambda}_{j-1}^T\bm{\varphi}_{\ell_1'}(\bm \xi_k)), \enskip \bm u^{(j)}_k := \bm u_{s}(\bm{\lambda}_{j}^T\bm{\varphi}_{\ell_2'}(\bm \xi_k)), \enskip \bm u^{(j+1)}_k := \bm u_{s}(\bm{\lambda}_{j+1}^T\bm{\varphi}_{\ell_3'}(\bm \xi_k)).
\end{align*}
Then, the finite volume update function \eqref{eq:adaptiveFVUpdate} can be written as
\begin{align}\label{eq:momentUpQuadrature}
\bm{c}_{\ell}^{\bm{\ell}'}&\left(\bm{\lambda}_{1},\bm{\lambda}_2,\bm{\lambda}_3\right)=\sum_{k=1}^Q w_k \left[\bm u^{(j)}_k- \frac{\Delta t}{\Delta x}\left(\bm g( \bm u^{(j)}_k,\bm u^{(j+1)}_k )- \bm g( \bm u^{(j-1)}_k,\bm u^{(j)}_k )\right)\right]\bm{\varphi}_{\ell}(\bm \xi_k)^T.
\end{align}
Instead of distributing the spatial mesh on the different processors, we now distribute the quadrature set, i.e. the sum in \eqref{eq:momentUpQuadrature} can be computed in parallel. Now after having performed the dual update, the dual variables are send to all processors. With these variables, each processor computes the solution on its portion of the quadrature set and then computes its part of the sum in \eqref{eq:momentUpQuadrature} on all spacial cells. All parts from the different processors are then added together and the full time-updated moments are distributed to all processors. From here, the dual update can again be performed. The standard IPM Algorithm~\ref{alg:IPM} and One-Shot IPM Algorithm~\ref{alg:osIPM} use this parallelization strategy accordingly. Again, we point out that stochastic-Galerkin is a variant of IPM, i.e. all presented techniques for IPM can also be used for SG. The SC algorithm that we use to compare intrusive with non-intrusive methods uses a given deterministic solver as a black box. Here, we distribute the quadrature set between all processors. Note that both, SC and IPM are based on the same deterministic solver, i.e. we use the same deterministic numerical flux $\bm g$. To our best knowledge, this allows a fair comparison of the different intrusive and non-intrusive techniques.

### 2019-11-25 Paper1/introduction.tex
last comments from Martin

When studying hyperbolic equations, the moment approximations of various methods such as Stochastic Galerkin \cite{le2004uncertainty}, IPM \cite{kusch2018filtered,poette2019contribution} and stochastic-Collocation \cite{barth2013non,dwight2013adaptive} tend to show incorrect discontinuities in certain regions of the physical space. These non-physical structures dissolve when the number of basis functions is increased \cite{pettersson2009numerical,offner2017stability} or when artificial diffusion is added through either the spatial numerical method \cite{offner2017stability} or by filters \cite{kusch2018filtered}. Also, a multi-element approach which divides the uncertain domain into cells and uses piece-wise polynomial basis functions to represent the solution has proven to mitigate non-physical discontinuities \cite{wan2006multi,durrwachter2018hyperbolicity}. Discontinuous structures commonly arise on a small portion of the space-time domain. Therefore, intrusive methods seem to be an adequate choice since they are well suited for adaptive strategies. By locally increasing the polynomial order \cite{tryoen2012adaptive,kroker2012finite,giesselmann2017posteriori} or adding artificial viscosity \cite{kusch2018filtered} at certain spatial positions and time steps in which complex structures such as discontinuities occur, a given accuracy can be reached with significantly reduced numerical costs. In addition to that, the number of moments needed to obtain a certain order with intrusive methods is smaller than the number of quadrature points for SC. An additional downside of collocation methods are aliasing effects, which stem from the inexact approximation of integrals. Consequently, collocation methods typically require a higher number of unknowns than intrusive methods to reach a given accuracy \cite{xiu2009fast,alekseev2011estimation}. Therefore, one aim should be to accelerate intrusive methods, since they can potentially outperform non-intrusive methods in complex and high-dimensional settings. \\

@@

The effectiveness of these acceleration ideas are tested by comparing results with stochastic-Collocation for the uncertain NACA test case as well as a bent shock tube problem. Our numerical studies show the following main results:
\begin{itemize}
\item In our test cases, the need to solve an optimization problem when using the IPM method leads to a significantly higher run time than SC and SG. However when using the discussed acceleration techniques, IPM requires the shortest time to reach a given accuracy.
\item Comparing SG with IPM, one observes that for the same number of unknowns, SG yields more accurate expectation values, whereas IPM shows improved variance approximations.
\item By studying aliasing effects, we show that SC requires a higher number of unknowns than intrusive methods (even for a one-dimension uncertainty) to reach the same accuracy level.
\item Using sparse grids for the IPM discretization when the space of uncertainty is multi-dimensional, the number of quadrature points needed to guarantee sufficient regularity of the Hessian matrix is significantly increased.
\end{itemize}
The IPM and SG calculations use a semi-intrusive numerical method, meaning that the discretization allows recycling a given deterministic code to generate the IPM solver. While facilitating the task of implementing general intrusive methods, this framework reduces the number of operations required to compute numerical fluxes. Also, it provides the ability to base the intrusive method on the same deterministic solver as the implementation used in a black-box fashion for stochastic-Collocation, which we believe allows a fair comparison between intrusive and non-intrusive methods. The code is publicly available to allow reproducibility \comment{[RefCode]}.

>>>>

When studying hyperbolic equations, the moment approximations of various methods such as Stochastic Galerkin \cite{le2004uncertainty}, IPM \cite{kusch2018filtered,poette2019contribution} and stochastic-Collocation \cite{barth2013non,dwight2013adaptive} tend to show incorrect discontinuities in certain regions of the physical space. These non-physical structures dissolve when the number of basis functions is increased \cite{pettersson2009numerical,offner2017stability} or when artificial diffusion is added through either the spatial numerical method \cite{offner2017stability} or by filters \cite{kusch2018filtered}. Also, a multi-element approach which divides the uncertain domain into cells and uses piece-wise polynomial basis functions to represent the solution has proven to mitigate non-physical discontinuities \cite{wan2006multi,durrwachter2018hyperbolicity}. Non-intrusive Monte-Carlo methods \cite{mishra2012multi,mishra2012sparse,mishra2016numerical}, which randomly sample input unceratinties to compute quantities of interest are robust, but suffer from a slow rate of convergence while again lacking the ability to use adaptivity to their full extend. Discontinuous structures commonly arise on a small portion of the space-time domain. Therefore, intrusive methods seem to be an adequate choice since they are well suited for adaptive strategies. By locally increasing the polynomial order \cite{tryoen2012adaptive,kroker2012finite,giesselmann2017posteriori} or adding artificial viscosity \cite{kusch2018filtered} at certain spatial positions and time steps in which complex structures such as discontinuities occur, a given accuracy can be reached with significantly reduced numerical costs. In addition to that, the number of moments needed to obtain a certain order with intrusive methods is smaller than the number of quadrature points for SC. An additional downside of collocation methods are aliasing effects, which stem from the inexact approximation of integrals. Consequently, collocation methods typically require a higher number of unknowns than intrusive methods to reach a given accuracy \cite{xiu2009fast,alekseev2011estimation}. Therefore, one aim should be to accelerate intrusive methods, since they can potentially outperform non-intrusive methods in complex and high-dimensional settings. \\

@@

The effectiveness of these acceleration ideas are tested by comparing results with stochastic-Collocation for the uncertain NACA test case as well as a bent shock tube problem. Our numerical studies show the following main results:
\begin{itemize}
\item In our test cases, the need to solve an optimization problem when using the IPM method leads to a significantly higher run time than SC and SG. However when using the discussed acceleration techniques, IPM requires the shortest time to reach a given accuracy.
\item Comparing SG with IPM, one observes that for the same number of unknowns, SG yields more accurate expectation values, whereas IPM shows improved variance approximations.
\item By studying aliasing effects, we show that SC requires a higher number of unknowns than intrusive methods (even for a one-dimension uncertainty) to reach the same accuracy level.
\item Using sparse grids for the IPM discretization when the space of uncertainty is multi-dimensional, the number of quadrature points needed to guarantee sufficient regularity of the Hessian matrix is significantly increased.
\end{itemize}
The IPM and SG calculations use a semi-intrusive numerical method, meaning that the discretization allows recycling a given deterministic code to generate the IPM solver. While facilitating the task of implementing general intrusive methods, this framework reduces the number of operations required to compute numerical fluxes. Also, it provides the ability to base the intrusive method on the same deterministic solver as the implementation used in a black-box fashion for stochastic-Collocation, which we believe allows a fair comparison between intrusive and non-intrusive methods. The code is publicly available to allow reproducibility \cite{uqcreator}.

### 2019-11-25 Paper1/summary_outlook.tex
last comments from Martin

\section{Summary and outlook}
\label{sec:summary_outlook}
In this work, we proposed acceleration techniques, which can be applied to the intrusive framework of IPM and SG: We use a one-shot technique to iterate moments and their corresponding dual variables to their steady state simultaneously. By not fully converging the dual iteration, we can reduce computational costs. Additionally, we make use of adaptivity and propose to keep a low maximal truncation order for most of the iteration to the steady state solution. Since complicated structures in the uncertain dimension only appear on a small portion of the spacial mesh, we are able to heavily reduce computational costs. The effects of the proposed techniques have been demonstrated by comparing results obtained with IPM as well as SG against SC. In our test cases, the intrusive methods yield the same error level as SC for a reduced runtime, especially since intrusive methods require less unknowns to achieve a certain accuracy due to aliasing errors. In higher-dimensional problems, this effect is amplified since the number of unknowns to achieve a certain total degree is asymptotically smaller than the number of quadrature points in a tensorized or sparse grid. Furthermore, we could observe that the required residual at which the solution of intrusive methods reaches a steady state is smaller than for SC. Additionally, the ability to adaptively change the truncation order helps intrusive methods to compete with SC in terms of computational runtime.

In future work, we aim at further accelerating the IPM method by using non-exact Hessian approximations. Similar to the one-shot idea of not fully converging the dual problem, it seems to be plausible to not spend too much time on computing the Hessian when the moments are not close to a steady state. Hessian approximations that can be interesting are BFGS and sparse BFGS \cite[Chapter~6.1]{nocedal2006numerical}, which construct the Hessian from previously computed gradients. Note that this strategy will increase the used memory, since old Hessians or gradients from a certain number of old time steps need to be saved in every spacial cell.
Even though the non-intrusive nature of stochastic-Collocation or Monte Carlo methods allows an easy implementation, it can be important to intrusively modify the code in order to fully exploit all acceleration potentials. Synchronizing the time updates of the solution at different quadrature points yields an increased control over the solution during the computation, which can for example be uses to employ adaptive methods. In this case one can switch to a fine quadrature level in a certain spatial cell by for example computing moments with the given coarse set of collocation points. From these moments one can compute an IPM reconstruction, which one can then evaluate at a finer quadrature set. Another example of breaking up the non-intrusive nature of Monte Carlo methods can be found in \cite{poette2019gpc}, where the generation of random samples is combined with the sampling after collisions to increase efficiency. Furthermore, we aim at applying the proposed acceleration techniques to SG methods with hyperbolicity limiters \cite{wu2017stochastic,schlachter2018hyperbolicity}.

>>>>

\section{Summary and outlook}
\label{sec:summary_outlook}
In this work, we proposed acceleration techniques, which can be applied to the intrusive framework of IPM and SG: We use a One-Shot technique to iterate moments and their corresponding dual variables to their steady state simultaneously. By not fully converging the dual iteration, we can reduce computational costs. Additionally, we make use of adaptivity and propose to keep a low maximal truncation order for most of the iteration to the steady state solution. Since complicated structures in the uncertain dimension only appear on a small portion of the spacial mesh, we are able to heavily reduce computational costs. The effects of the proposed techniques have been demonstrated by comparing results obtained with IPM as well as SG against SC. In our test cases, the intrusive methods yield the same error level as SC for a reduced runtime, especially since intrusive methods require less unknowns to achieve a certain accuracy due to aliasing errors. In higher-dimensional problems, this effect is amplified since the number of unknowns to achieve a certain total degree is asymptotically smaller than the number of quadrature points in a tensorized or sparse grid (\eqref{eq:numberBasisFcts} instead of $O(M(\log_2(M)^{p-1}))$). Furthermore, we could observe that the required residual at which the solution of intrusive methods reaches a steady state is smaller than for SC. Additionally, the ability to adaptively change the truncation order helps intrusive methods to compete with SC in terms of computational runtime.

In future work, we aim at further accelerating the IPM method by using non-exact Hessian approximations. Similar to the One-Shot idea of not fully converging the dual problem, it seems to be plausible to not spend too much time on computing the Hessian when the moments are not close to a steady state. Hessian approximations that can be interesting are BFGS and sparse BFGS \cite[Chapter~6.1]{nocedal2006numerical}, which construct the Hessian from previously computed gradients. Note that this strategy will increase the used memory, since old Hessians or gradients from a certain number of old time steps need to be saved in every spacial cell.
Even though the non-intrusive nature of stochastic-Collocation or Monte Carlo methods allows an easy implementation, it can be important to intrusively modify the code in order to fully exploit all acceleration potentials. Synchronizing the time updates of the solution at different quadrature points yields an increased control over the solution during the computation, which can for example be uses to employ adaptive methods. In this case one can switch to a fine quadrature level in a certain spatial cell by for example computing moments with the given coarse set of collocation points. From these moments one can compute an IPM reconstruction, which one can then evaluate at a finer quadrature set. Another example of breaking up the non-intrusive nature of Monte Carlo methods can be found in \cite{poette2019gpc}, where the generation of random samples is combined with the sampling after collisions to increase efficiency. Furthermore, we aim at applying the proposed acceleration techniques to SG methods with hyperbolicity limiters \cite{wu2017stochastic,schlachter2018hyperbolicity}.
